---
title: "Web application with sales forecasting"
output: html_document
prepaer: Dawood Afzal
runtime: shiny
---

```{r echo = FALSE,, warning=FALSE, message=FALSE}
library(flexdashboard)
library(tidyverse)
library(forecast)
library(shiny)
library(dplyr)
library(ggplot2)
library(DT)
```


<!--- ## Importing libraries and setting up a basic ARIMA forecast mode --->

<!--- ### The following code is for test purposes in the report only. Note later on- I utilized some of the tables generated from the starter code and leaving this code with results not displayed. 

In the following code, all libraries used will be imported along with reading the CSV data source file from Github.  
The following steps have been applied:  
1. Split the 142 weeks sales history into 129 weeks for training the ARIMA model (sales_hist) and 13 weeks (sales_last) for forecast comparison.  
2. Plotting the sales_hist table in a time series, segregating years = 52 weeks.  
3. Creating a new data frame with columns predicted (ARIMA model forecast representing 13 weeks or a quarter), actual(representing actual sales for last 13 weeks). This can be plotted to test the forecast vs actual quarter.  
4. Lastly and entire time series for actual sales is generated.  --->



```{r, echo = FALSE, warning=FALSE, message=FALSE, include = FALSE}


# historical records from multiple stores
fsales <- "https://raw.githubusercontent.com/multidis/hult-inter-bus-reports-r/main/forecasting/sales_weekly.csv"
sales <- read_csv(fsales)

# latest (current) week
nweek_now <- max(sales$Week)

# sales in store 3 over the most recent quarter;
# counting 52 weeks/year, 13 weeks/quarter
sales %>%
  filter(Store == 3) %>%
  subset(Week > nweek_now - 13, Week <= nweek_now) %>%
  summarise(TotalSales = sum(Weekly_Sales))

# check forecast accuracy for the most recent quarter
sales_hist <- sales %>%
  filter(Store == 3) %>%
  subset(Week <= nweek_now - 13)
sales_last <- sales %>%
  filter(Store == 3) %>%
  subset(Week > nweek_now - 13)
```

<!--- ####      Historical Sales for Store 3 --->

```{r, echo = FALSE, warning=FALSE, message=FALSE, include = FALSE}
# time series with annual periodicity to account seasonality
sales_hist_ts <- ts(sales_hist$Weekly_Sales, frequency = 52)
autoplot(sales_hist_ts)
labs(title = "Sales History")

# ARIMA: Auto-Regressive Integrated Moving Average
# methodological details:
#  https://otexts.com/fpp3/arima.html
arima_model <- auto.arima(sales_hist_ts, seasonal.test = "seas")

# forecast horizon to match most recent quarter
arima_pred <- forecast(arima_model, h = 13)

# note: confidence intervals (lower, upper) are available as well
sales_pred_eval <- data.frame(predicted = as.numeric(arima_pred$mean),
                              actual = sales_last$Weekly_Sales,
                              Week = sales_last$Week)
sales_pred_eval %>%
  ggplot(aes(x = Week)) +
  geom_line(aes(y = predicted, col = "red")) +
  geom_line(aes(y = actual, col = "green")) +
  ggtitle("Store 3 | Sales forecast vs Actual Sales | Quarter")

```

<!--- ####      Time series for actual sales for Store 3 --->

```{r, echo = FALSE, warning=FALSE, message=FALSE, include = FALSE}
# time series for the entire dataset to use for future forecasting
sales_ts <- ts(sales %>% filter(Store == 3) %>% select(Weekly_Sales), frequency = 52)
autoplot(sales_ts)

# forecasting for the next 2 months
arima_model <- auto.arima(sales_ts, seasonal.test = "seas")
arima_pred <- forecast(arima_model, h = 8)
arima_pred$mean



```



## 1. Actual vs Forecast Timeseries for each Store with period selection
Select Store from the dropdown menu.  
Select forecast period to display

```{r echo = FALSE}
#utilized the input panel from the starter code with a max value of 52 to show 52 weeks of forecast, 

inputPanel(
  selectInput("storenum", label = "Select store: ",
              choices = 1:max(sales$Store), selected = 1),
  sliderInput("horizon", label = "Forecast horizon (weeks): ",
              min = 2, max = 52, value = 52, step = 1)
)
```

### Timeseries : Actual vs Forecast

```{r echo = FALSE}
#amended the reactive block to show prior_year as a comparison to forecast. Added geom_line code to differentiate the actual and forecast lines

renderPlot({
  horiz <- input$horizon
  
  sales_ts <- ts(sales %>% filter(Store == input$storenum) %>%
                   select(Weekly_Sales), frequency = 52)
  arima_model <- auto.arima(sales_ts, seasonal.test = "seas")
  arima_pred <- forecast(arima_model, h = horiz)
  
  sales_pred <- data.frame(Sales_Predicted = as.numeric(arima_pred$mean),
                           Week = (nweek_now + 1):(nweek_now + horiz))
  
  prior_year <- sales %>%
    filter(Store == input$storenum) %>%
    subset(Week <= 142)
  
  sales_pred %>% 
    ggplot() +
    geom_line(aes(Week, Sales_Predicted, color = "forecast"), linetype = "solid", size = 1) +
    geom_line(data = prior_year, aes(Week, Weekly_Sales, color = "PY Actuals"), linetype = "solid", size = 1) +
    labs(x = "Weeks", y = "Weekly Sales") +
    theme_minimal() +
    scale_color_manual(values = c("forecast" = "blue", "PY Actuals" = "red"),
                       labels = c("forecast", "PY Actuals"),
                       name = "Lines") +
    guides(color = guide_legend(override.aes = list(linetype = c("solid", "solid"), size = c(1, 1))))
})

```

## 2. Generate sales forecast for a given week (within the next three months) selected in the UI

## Solution

### Create a data table  for splitting weeks into three columns Y1 and Y2 (for actuals) and Y3 for forecast

Steps utilized:  
1. Created a table showing Y1, Y2 and Y3 (Forecast based on ARIMA model) by store through the same function as utilized before.  
2. Applied the function to create a list called results_list.  
4. Stored the list in a new table and printed table to manually check the values from original sales table. 


```{r echo = FALSE, , warning=FALSE, message=FALSE}
# Creating a function that can be repeated to calculate Y1, Y2, and Y3(forecast) for a given store
calculate_sales <- function(store_id) {
  # Here we extract the  sales data for the store by store id
  sales_store <- sales %>% filter(Store == store_id)
  
  # For the extracted store data above, create a time series for weekly sales by store, with annualized frequency
  sales_ts <- ts(sales_store$Weekly_Sales, frequency = 52)
  
  # Essentially the same step as in the starter code, create an arima model with seasonality
  arima_model <- auto.arima(sales_ts, seasonal.test = "seas")
  
  # Storing the forecast model for the next 52 weeks - 
  arima_pred <- forecast(arima_model, h = 52)
  
  # Splitting sales data by year utilizing the sequence of weeks in the stored data for actual sales
  sales_years <- split(sales_store$Weekly_Sales, ceiling(seq_along(sales_store$Weekly_Sales)/52))
  
  # Storing data in sales_years table in year 1, year 2 and storing data in Arima_pred forecast values in Y3
  y1 <- sales_years[[1]]
  y2 <- sales_years[[2]]
  y3 <- arima_pred$mean
  
  # Creating a data frame with store ID, month, and corresponding values
  store_data <- data.frame(Store = rep(store_id, length(y1)),
                           Week = seq_along(y1),
                           Y1 = y1,
                           Y2 = y2,
                           Y3 = y3)
  
  # Returning the results as a data frame
  return(store_data)
}

# Applying the function for each individual store to all stores and storing the results in a list
all_stores <- unique(sales$Store)
result_list <- lapply(all_stores, calculate_sales)

# Combining the results of the list in a single data frame for use and printing the first 10 rows
new_table <- bind_rows(result_list)
print(head(new_table, n = 10))

```

### Create an interactive plot and table for comparisons between same weeks of respective Y1 and Y2 (Actuals) and Y3 (Forecast)

```{r echo = FALSE, , warning=FALSE, message=FALSE}
# calling all relevant librarires


# UI code
ui <- fluidPage(
  titlePanel("Weekly Sales Comparison"),
  sidebarLayout(
    sidebarPanel(
      selectInput("store", "Select Store:", choices = unique(sales$Store)),
      actionButton("reset", "Reset Selection"),
      width = 2
    ),
    mainPanel(
      plotOutput("salesPlot", height = "500px"),
      dataTableOutput("salesTable"),
      width = 10
    )
  )
)

# Server code
server <- function(input, output, session) {
  # Function to calculate Y1, Y2, and Y3 for a given store
  calculate_sales <- function(store_id) {
    # Extracting sales data for the store
    sales_store <- sales %>% filter(Store == store_id)
    
    # Creating a time series for the store's sales data
    sales_ts <- ts(sales_store$Weekly_Sales, frequency = 52)
    
    # Fitting an ARIMA model to the time series
    arima_model <- auto.arima(sales_ts, seasonal.test = "seas")
    
    # Forecasting for the next year)
    arima_pred <- forecast(arima_model, h = 52)
    
    # Splitting sales data by year
    sales_years <- split(sales_store$Weekly_Sales, ceiling(seq_along(sales_store$Weekly_Sales)/52))
    
    # Storing sales values for each year in separate variables
    y1 <- sales_years[[1]]
    y2 <- sales_years[[2]]
    y3 <- arima_pred$mean
    
    # Creating a data frame with week and corresponding values
    sales_data <- data.frame(Week = 1:52,
                             Y1 = y1,
                             Y2 = y2,
                             Y3F = y3)
    
    # Returning the results as a data frame
    return(sales_data)
  }
  
  # Reactive function to update the sales table
  sales_table <- reactive({
    store_id <- input$store
    calculate_sales(store_id)
  })
  
  # Render the sales table
  output$salesTable <- DT::renderDataTable({
    sales_table()
  }, options = list(paging = FALSE))
  
  # Reactive function to update the sales plot
  sales_plot <- reactive({
    store_id <- input$store
    sales_data <- calculate_sales(store_id)
    
    # Melt the data for plotting
    sales_data_melted <- sales_data %>%
      gather("Year", "Sales", -Week)
    
    # Create the plot
    ggplot(sales_data_melted, aes(x = Week, y = Sales, color = Year)) +
      geom_line() +
      labs(x = "Week", y = "Sales", color = "Year")
  })
  
  # Render the sales plot
  output$salesPlot <- renderPlot({
    sales_plot()
  })
  
  # Reset the selection
  observeEvent(input$reset, {
    updateSelectInput(session, "store", selected = NULL)
  })
}

# Run the Shiny app
shinyApp(ui = ui, server = server)

```


## 3. Incorporate a UI functionality to compare monthly sales between the past year (actual) and the upcoming year (forecast).

## Solution

### 1. Create a table with monthly aggregation for sales for Y1, Y2 and Y3 (forecast)

Steps utilized:  
1. Created a table showing Y1, Y2 and Y3(Forecast based on ARIMA model) by store through the same function as utilized before.  
2. Applied the function to create a list called results_list.  
3. Utilized the list to sum up sales for every 4 weeks and store in a table.  
4. Printed table to manually check the values from original sales table. 


```{r echo = FALSE, message = FALSE}
# Function to calculate Y1, Y2, and Y3 for a given store, same function as before utilized
calculate_sales <- function(store_id) {
  # Extracting sales data for the store
  sales_store <- sales %>% filter(Store == store_id)
  
  # For the extracted store data above, create a time series for weekly sales by store, with annualized frequncy
  sales_ts <- ts(sales_store$Weekly_Sales, frequency = 52)
  
  # Essentially the same step as in the starter code, create an arima model with seasonality
  arima_model <- auto.arima(sales_ts, seasonal.test = "seas")
  
  # Storing the forecast model for the next 52 weeks - I have utilized 52 weeks in monthly ARIMA model as well to ensure seasonality gets captured
  arima_pred <- forecast(arima_model, h = 52)
  
  # Splitting sales data by year utilizing the sequence of weeks in the stored data for actual sales
  sales_years <- split(sales_store$Weekly_Sales, ceiling(seq_along(sales_store$Weekly_Sales)/52))
  
  # Storing data in sales_years table in year 1, year 2 and storing data in Arima_pred forecast values in Y3
  y1 <- sales_years[[1]]
  y2 <- sales_years[[2]]
  y3 <- arima_pred$mean
  
  # Creating a data frame with store ID, month, and corresponding values
  store_data <- data.frame(Store = rep(store_id, length(y1)),
                           Week = seq_along(y1),
                           Y1 = y1,
                           Y2 = y2,
                           Y3F = y3)
  
  # Returning the results as a data frame
  return(store_data)
}

# Applying the function to all stores and storing the results in a list, the results list compiles each store by week, Y1 sales, Y2 sales and Y3 forecast sales
all_stores <- unique(sales$Store)
result_list <- lapply(all_stores, calculate_sales)

# Combining the results into a single data frame, with bind_rows used to sort the vectors vertically, and monthly table is created by grouping by month at every 4th week. This is an interesting use indicing and contrary to diving week by 4 as it looks, combined with group_by and summarize, it sums up every 4 weeks.
monthly_table <- bind_rows(result_list)
monthly_table <- monthly_table %>%
  group_by(Store, Month = ceiling(Week / 4)) %>%
  summarise(Y1 = sum(Y1),
            Y2 = sum(Y2),
            Y3F = sum(Y3F)) %>%
  ungroup()

#printing the table to check if the results are correct through manual verification
print(head(monthly_table, n = 10))


```


### 2. Generate bar plot comparing monthly sales YOY and display interactive inputs for store and month

```{r echo = FALSE}

# Function to calculate monthly sales for a selected store and month
calculate_monthly_sales <- function(store_id, selected_month) {
  # Subset the monthly_table based on the selected store and month, filtering the data from the month_table created before 
  filtered_data <- monthly_table %>% filter(Store == store_id, Month == selected_month)
  
  # Return the filtered data
  return(filtered_data)
}

# UI code - utilizing fluidPage in Shiny to create UI interface, with a sidebar Panel for store and month selection. Main panel is used to display barplot and table together.
ui <- fluidPage(
  titlePanel("Monthly Sales Comparison"),
  sidebarLayout(
    sidebarPanel(
      selectInput("store", "Select a store:", choices = unique(monthly_table$Store)),
      selectInput("month", "Select a month:", choices = unique(monthly_table$Month))
    ),
    mainPanel(
      tableOutput("sales_table"),
      plotOutput("sales_plot")
    )
  )
)

# Server code
server <- function(input, output) {
  # Render the sales table
  output$sales_table <- renderTable({
    # Get the selected store and month, linking the input function
    store_id <- input$store
    selected_month <- input$month
    
    # Calculate the monthly sales for the selected store and month
    sales_data <- calculate_monthly_sales(store_id, selected_month)
    
    # Return the sales table
    sales_data
  })
  
  # Render the sales plot with reactive inputs connected
  output$sales_plot <- renderPlot({
    # Get the selected store and month - same way linking the input function
    store_id <- input$store
    selected_month <- input$month
    
    # Calculate the monthly sales for the selected store and month using the function created above
    sales_data <- calculate_monthly_sales(store_id, selected_month)
    
    # Reshaped the data from wide to long format - the pivot longer i'm not sure but it works to not overlap the chart with the table
    sales_data_long <- sales_data %>%
      pivot_longer(cols = c(Y1, Y2, Y3F), names_to = "Year", values_to = "Sales")
    
    # Plot the sales comparison - standard ggplot barplot 
    ggplot(sales_data_long, aes(x = Year, y = Sales, fill = Year)) +
      geom_col(width = 0.5) +
      labs(x = "Year", y = "Sales", title = "Monthly Sales Comparison") +
      theme_minimal()
  })
}

# Run the Shiny app
shinyApp(ui = ui, server = server)



```

## 4. Generate sales forecast for a specific future quarter (during the upcoming year) selected in the UI

## Solution

### 1. Create a table with quarterly aggregation for sales for Y1, Y2 and Y3 (forecast)

Steps utilized:  
1. Created a table showing Y1, Y2 and Y3(Forecast based on ARIMA model) by store through the same function as utilized before.  
2. Applied the function to create a list called results_list.  
3. Utilized the list to sum up sales for every 13 weeks and store in a table.  
4. Printed table to manually check the values from original sales table. 


```{r echo = FALSE, message = FALSE}
# Function to calculate Y1, Y2, and Y3 for a given store, same function as before utilized
calculate_sales <- function(store_id) {
  # Extracting sales data for the store
  sales_store <- sales %>% filter(Store == store_id)
  
  # For the extracted store data above, create a time series for weekly sales by store, with annualized frequency
  sales_ts <- ts(sales_store$Weekly_Sales, frequency = 52)
  
  # Essentially the same step as in the starter code, create an arima model with seasonality
  arima_model <- auto.arima(sales_ts, seasonal.test = "seas")
  
  # Storing the forecast model for the next 52 weeks - I have utilized 52 weeks in monthly ARIMA model as well to ensure seasonality gets captured
  arima_pred <- forecast(arima_model, h = 52)
  
  # Splitting sales data by year utilizing the sequence of weeks in the stored data for actual sales
  sales_years <- split(sales_store$Weekly_Sales, ceiling(seq_along(sales_store$Weekly_Sales) / 52))
  
  # Storing data in sales_years table in year 1, year 2, and storing data in Arima_pred forecast values in Y3
  y1 <- sales_years[[1]]
  y2 <- sales_years[[2]]
  y3 <- arima_pred$mean
  
  # Creating a data frame with store ID, week, and corresponding values
  store_data <- data.frame(Store = rep(store_id, length(y1)),
                           Week = seq_along(y1),
                           Y1 = y1,
                           Y2 = y2,
                           Y3F = y3)
  
  # Returning the results as a data frame
  return(store_data)
}

# Applying the function to all stores and storing the results in a list
all_stores <- unique(sales$Store)
result_list <- lapply(all_stores, calculate_sales)

# Combining the results into a single data frame using bind_rows
quarterly_table <- bind_rows(result_list)

# Grouping the quarterly_table by Store and creating the Month variable based on Week
quarterly_table <- quarterly_table %>%
  group_by(Store, Quarter = ceiling(Week / 13)) %>%
  summarise(Y1 = sum(Y1),
            Y2 = sum(Y2),
            Y3F = sum(Y3F)) %>%
  ungroup()

# Printing the table to check the results
print(head(quarterly_table, n = 10))


```

### 2. Generate bar plot comparing quarterly sales YOY and display interactive inputs for store and quarter selection

```{r echo = FALSE}

# Function to calculate quarterly sales for a selected store and quarter
calculate_quarterly_sales <- function(store_id, selected_quarter) {
  # Subset the quarterly_table based on the selected store and quarter, filtering the data from the quarterly_table created before 
  filtered_data <- quarterly_table %>% filter(Store == store_id, Quarter == selected_quarter)
  
  # Return the filtered data
  return(filtered_data)
}

# UI code - utilizing fluidPage in Shiny to create UI interface, with a sidebar Panel for store and quarter selection. Main panel is used to display barplot and table together.
ui <- fluidPage(
  titlePanel("Quarterly Sales Comparison"),
  sidebarLayout(
    sidebarPanel(
      selectInput("store", "Select a store:", choices = unique(quarterly_table$Store)),
      selectInput("quarter", "Select a quarter:", choices = unique(quarterly_table$Quarter))
    ),
    mainPanel(
      tableOutput("sales_table"),
      plotOutput("sales_plot")
    )
  )
)

# Server code
server <- function(input, output) {
  # Render the sales table
  output$sales_table <- renderTable({
    # Get the selected store and quarter, linking the input function
    store_id <- input$store
    selected_quarter <- input$quarter
    
    # Calculate the quarterly sales for the selected store and quarter
    sales_data <- calculate_quarterly_sales(store_id, selected_quarter)
    
    # Return the sales table
    sales_data
  })
  
  # Render the sales plot with reactive inputs connected
  output$sales_plot <- renderPlot({
    # Get the selected store and quarter - same way linking the input function
    store_id <- input$store
    selected_quarter <- input$quarter
    
    # Calculate the quarterly sales for the selected store and quarter using the function created above
    sales_data <- calculate_quarterly_sales(store_id, selected_quarter)
    
    # Reshape the data from wide to long format- the pivot longer i'm not sure but it works to not overlap the chart with the table
    sales_data_long <- sales_data %>%
      pivot_longer(cols = c(Y1, Y2, Y3F), names_to = "Year", values_to = "Sales")
    
    # Plot the sales comparison - standard ggplot barplot
    ggplot(sales_data_long, aes(x = Year, y = Sales, fill = Year)) +
      geom_col(width = 0.5) +
      labs(x = "Year", y = "Sales", title = "Quarterly Sales Comparison") +
      theme_minimal()
  })
}

# Run the Shiny app
shinyApp(ui = ui, server = server)



```

